---
title: "Project 4: Movie Recommendation Report by ZM11"
author: Zhouning Ma (ZM11)
date: '09/07/2020'
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan

fontfamily: mathpazo
fontfamilyoptions: sc, osf
fontsize: 11pt
linkcolor: blue
---

***



```{r include=FALSE}
start_time <- Sys.time()

library(recommenderlab)
library(ggplot2)
library(data.table)
library(reshape2)
library(knitr)
```


### Project Overview:
In this project, we develop content based recommendation system(System I) and collaborative recommendation system(System II) (1).

* **System I:** recommendation based on genres. We know user's favorite genre, we provide two recommendation schemes: 
   + __Method I:__  highly-rate movies - select the 4-point or more review
   + __Method II:__ most trendy movies - the newest movie(here is year 2000)

* **System II:** collaborative recommendation system. we provided three collaborative recommendation algorithms: 
   + __User-based (UBCF):__ we assume that similar users will have similar taste. UBCF uses the logic and recommends items by finding similar users to the user;
   + __Item-based (IBCF):__ The similarities between different items in the dataset are calculated by using one of a number of similarity measures, and then these similarity values are used to predict;
   + __Singular value decomposition (SVD):__ It uses a matrix structure where each row represents a user, and each column represents an item. The elements of this matrix are the ratings that are given to items by users.


```{r include=FALSE}
read <- function(fileName, separators) {
    data <- readLines(con <- file(fileName))
    close(con)
    records <- sapply(data, strsplit, split=separators)
    dataFrame <- data.frame(t(sapply(records,c)))
    rownames(dataFrame) <- 1: nrow(dataFrame)
    return(as.data.frame(dataFrame,stringsAsFactors = FALSE))
}


basedir ="data/test/"
movies = read(paste0(basedir,"movies.dat"), "::")
ratings = read(paste0(basedir,"ratings.dat"), "::")
users = read(paste0(basedir, "users.dat"), "::")
colnames(movies) = c('MovieID', 'title', 'genres')
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

```


### Dataset: 
The dataset(2) contains about 1 million ratings and the details as following:

* Number of movies: $`r nrow(movies)`$
* Number of ratings: $`r nrow(ratings)`$
* Number of users: $`r nrow(users)`$

The dataset README(3) shows:
Genres are pipe-separated and are selected from the following genres:

* Action
* Adventure
* Animation
* Children's
* Comedy
* Crime
* Documentary
* Drama
* Fantasy
* Film-Noir
- Horror
- Musical
- Mystery
- Romance
- Sci-Fi
- Thriller
- War
- Western
	



```{r include=FALSE}

## Data pre-processing
genres <- as.data.frame(movies[,3], stringsAsFactors=FALSE)
genres2 <- as.data.frame(tstrsplit(genres[,1], '[|]', type.convert=TRUE), stringsAsFactors=FALSE)
colnames(genres2)<- c(1:5)  #c(1:10)

genre_list <- c("Action", "Adventure", "Animation", "Children", 
                "Comedy", "Crime","Documentary", "Drama", "Fantasy",
                "Film-Noir", "Horror", "Musical", "Mystery","Romance",
                "Sci-Fi", "Thriller", "War", "Western")


genre_matrix <- matrix(0,nrow(movies) + 1,18) #empty matrix, 18=no of genres
genre_matrix[1,] <- genre_list #set first row to genre list
colnames(genre_matrix) <- genre_list #set column names to genre list

#iterate through matrix
for (i in 1:nrow(genres2)) {
  for (c in 1:ncol(genres2)) {
    genmat_col = which(genre_matrix[1,] == genres2[i,c])
    genre_matrix[i+1,genmat_col] <- 1
  }
}

#convert into dataframe
genre_matrix2 <- as.data.frame(genre_matrix[-1,], stringsAsFactors=FALSE) #remove first row, which was the genre list
for (c in 1:ncol(genre_matrix2)) {
  genre_matrix2[,c] <- as.integer(genre_matrix2[,c])
} 

#Create a matrix to search for a movie by genre:
years <- as.data.frame(movies$title, stringsAsFactors=FALSE)

substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

years <- as.data.frame(substr(substrRight(substrRight(years$`movies$title`, 6),5),1,4))
colMin <- function(data) sapply(data, min, na.rm = TRUE)
colMax <- function(data) sapply(data, max, na.rm = TRUE)

min_year = colMin(years)
max_year = colMax(years)

```

The movies' year are between $`r min_year`$ to $`r max_year`$.


```{r include=FALSE}
## Create a user profile
binaryratings <- ratings

# ratings of 4 and 5 are mapped to 1, 
# representing likes, and ratings of 3 
# and below are mapped to -1, representing 
# dislikes:

for (i in 1:nrow(binaryratings)){
  if (binaryratings[i,3] > 3){
    binaryratings[i,3] <- 1
  }
  else{
    binaryratings[i,3] <- -1
  }
}

# convert binaryratings matrix to the correct format:
binaryratings2 <- dcast(binaryratings, MovieID~UserID, value.var = "Rating", na.rm=FALSE)
#head(binaryratings2)


for (i in 1:ncol(binaryratings2)){
  binaryratings2[which(is.na(binaryratings2[,i]) == TRUE),i] <- 0
}
binaryratings2 = binaryratings2[,-1] #remove MovieIDs col. Rows are MovieIDs, cols are userIds
#head(binaryratings2)


#Remove rows that are not rated from movies dataset
MovieIDs <- length(unique(movies$MovieID)) #10329
ratingMovieIDs <- length(unique(ratings$MovieID)) #10325
movies2 <- movies[-which((movies$MovieID %in% ratings$MovieID) == FALSE),]
rownames(movies2) <- NULL

#Remove rows that are not rated from genre_matrix2
genre_matrix3 <- genre_matrix2[-which((movies$MovieID %in% ratings$MovieID) == FALSE),]
rownames(genre_matrix3) <- NULL

#Calculate dot product for User Profiles
result = matrix(0,18,nrow(users)) 
for (c in 1:ncol(binaryratings2)){
  for (i in 1:ncol(genre_matrix3)){
    result[i,c] <- sum((genre_matrix3[,i]) * (as.numeric(binaryratings2[,c]))) #ratings per genre
  }
}

#Convert to Binary scale
for (c in 1:ncol(result)){
  for (i in 1:nrow(result)){
    if (result[i,c] < 0){
      result[i,c] <- 0
    }
    else {
      result[i,c] <- 1
    }
  }
}

## Assume that users like similar items, and retrieve movies 
# that are closest in similarity to a user's profile, which 
# represents a user's preference for an item's feature.
# use Jaccard Distance to measure the similarity between user profiles

# The User-Based Collaborative Filtering Approach

#Create ratings matrix. Rows = userId, Columns = movieId
ratingmat <- dcast(ratings, UserID~MovieID, value.var = "Rating", na.rm=FALSE)
ratingmat <- as.matrix(ratingmat[,-1]) #remove userIds

# Method: UBCF
# Similarity Calculation Method: Cosine Similarity
# Nearest Neighbors: 30

#Convert rating matrix into a recommenderlab sparse matrix
ratingmat <- as(ratingmat, "realRatingMatrix")

# Determine how similar the first four users are with each other
# create similarity matrix
similarity_users <- similarity(ratingmat[1:4, ], 
                               method = "cosine", 
                               which = "users")
as.matrix(similarity_users)

```






### Exploring Similarity Data

```{r echo=TRUE}

image(as.matrix(similarity_users), main = "User similarity")

```


```{r include=FALSE}


# compute similarity between
# the first four movies
similarity_items <- similarity(ratingmat[, 1:4], method =
                                 "cosine", which = "items")
as.matrix(similarity_items)

```


```{r echo=TRUE}

image(as.matrix(similarity_items), main = "Item similarity")

```


### Most Viewed Movies Visualization (4)(5): 

```{r include=FALSE}

# Exploring values of ratings:
vector_ratings <- as.vector(ratingmat@data)
unique(vector_ratings) # what are unique values of ratings

table_ratings <- table(vector_ratings) # what is the count of each rating value
table_ratings


```


```{r echo=TRUE}
# Visualize the rating:
vector_ratings <- vector_ratings[vector_ratings != 0] # rating == 0 are NA values
vector_ratings <- factor(vector_ratings)

qplot(vector_ratings) + 
  ggtitle("Distribution of the ratings")

```


```{r echo=TRUE}


# Exploring viewings of movies:
views_per_movie <- colCounts(ratingmat) # count views for each movie
table_views <- data.frame(movie = names(views_per_movie), views = views_per_movie) # create dataframe of views
table_views <- table_views[order(table_views$views, decreasing = TRUE), ] # sort by number of views


```


```{r echo=TRUE}

#head(movies2)

ggplot(table_views[1:6, ], aes(x = movie, y = views)) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_x_discrete(labels=subset(movies2, movies2$MovieID == table_views$movie)$title) +
  ggtitle("Number of views of the top movies")

#Visualizing the matrix:
image(ratingmat, main = "Heatmap of the rating matrix") # hard to read-too many dimensions
image(ratingmat[1:10, 1:15], main = "Heatmap of the first rows and columns")
image(ratingmat[rowCounts(ratingmat) > quantile(rowCounts(ratingmat), 0.99),
                 colCounts(ratingmat) > quantile(colCounts(ratingmat), 0.99)], 
      main = "Heatmap of the top users and movies")


#Normalize the data
ratingmat_norm <- normalize(ratingmat)
image(ratingmat_norm[rowCounts(ratingmat_norm) > quantile(rowCounts(ratingmat_norm), 0.99),
                colCounts(ratingmat_norm) > quantile(colCounts(ratingmat_norm), 0.99)], 
      main = "Heatmap of the top users and movies")



```


## Create realRatingMatrix

```{r}
ratings <- as(ratings, 'realRatingMatrix')
```


## Training and Testing Data
We split the data into 80% training and 20% testing. 

```{r}
train_proportion <- 0.8
to_keep <- 15  ## given 15 items
threshold <- 0 ## ratings above 0 as the cutoff point

# split the data into the training and the test set:
e <- evaluationScheme(ratings, method="split", train=train_proportion, k=10, given=to_keep, goodRating=threshold)
e

```


## Train User-based and Item-based Collaborative Filtering models

We will train CF models by difference parameters (7,8):

- normalize: NULL, center or Z-score;

- method: Cosine, Euclidean pearson

also, we will train two SVD models with difference parameter, total **20 models:**, we will calculate prediction accuracy as well, the following are the examples:


### User-Based Collaborative Filtering: Cosine Similarity

   + UBCF_N_C : The raw data is used with no normalization applied;

   + UBCF_C_C: Data are normalized using centering.

   + UBCF_Z_C: Z-score normalization is applied to the data;

```{r}
# train UBCF cosine similarity models
# non-normalized
UBCF_N_C <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = NULL, method="Cosine"))

# centered
UBCF_C_C <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "center",method="Cosine"))

# Z-score normalization
UBCF_Z_C <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "Z-score",method="Cosine"))

```

### Evaluate the models:

```{r}
# compute predicted ratings
p1 <- predict(UBCF_N_C, getData(e, "known"), type="ratings")
p2 <- predict(UBCF_C_C, getData(e, "known"), type="ratings")
p3 <- predict(UBCF_Z_C, getData(e, "known"), type="ratings")

# aggregate the performance statistics
error_UCOS <- rbind(
  UBCF_N_C = calcPredictionAccuracy(p1, getData(e, "unknown")),
  UBCF_C_C = calcPredictionAccuracy(p2, getData(e, "unknown")),
  UBCF_Z_C = calcPredictionAccuracy(p3, getData(e, "unknown"))
)
kable(error_UCOS)
```


### User-Based Collaborative Filtering: Euclidean Distance

```{r}
# train UBCF Euclidean Distance models

# non-normalized
UBCF_N_E <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = NULL, method="Euclidean"))

# centered
UBCF_C_E <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "center",method="Euclidean"))

# Z-score normalization
UBCF_Z_E <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "Z-score",method="Euclidean"))

```



```{r}
# compute predicted ratings
p1 <- predict(UBCF_N_E, getData(e, "known"), type="ratings")
p2 <- predict(UBCF_C_E, getData(e, "known"), type="ratings")
p3 <- predict(UBCF_Z_E, getData(e, "known"), type="ratings")

# aggregate the performance statistics
error_UEUC <- rbind(
  UBCF_N_E = calcPredictionAccuracy(p1, getData(e, "unknown")),
  UBCF_C_E = calcPredictionAccuracy(p2, getData(e, "unknown")),
  UBCF_Z_E = calcPredictionAccuracy(p3, getData(e, "unknown"))
)
kable(error_UEUC)
```

### User-Based Collaborative Filtering: Pearson Correlation

```{r}
#train UBCF pearson correlation models

# non-normalized
UBCF_N_P <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = NULL, method="pearson"))

# centered
UBCF_C_P <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "center",method="pearson"))

# Z-score normalization
UBCF_Z_P <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "Z-score",method="pearson"))

```


```{r}
# compute predicted ratings
p1 <- predict(UBCF_N_P, getData(e, "known"), type="ratings")
p2 <- predict(UBCF_C_P, getData(e, "known"), type="ratings")
p3 <- predict(UBCF_Z_P, getData(e, "known"), type="ratings")

# aggregate the performance statistics
error_UPC <- rbind(
  UBCF_N_P = calcPredictionAccuracy(p1, getData(e, "unknown")),
  UBCF_C_P = calcPredictionAccuracy(p2, getData(e, "unknown")),
  UBCF_Z_P = calcPredictionAccuracy(p3, getData(e, "unknown"))
)
kable(error_UPC)
```


### Item-Based Collaborative Filtering: Cosine Similarity

```{r}
#train IBCF cosine similarity models

# non-normalized
IBCF_N_C <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = NULL, method="Cosine"))

# centered
IBCF_C_C <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "center",method="Cosine"))

# Z-score normalization
IBCF_Z_C <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "Z-score",method="Cosine"))

```


```{r}
# compute predicted ratings
p1 <- predict(IBCF_N_C, getData(e, "known"), type="ratings")
p2 <- predict(IBCF_C_C, getData(e, "known"), type="ratings")
p3 <- predict(IBCF_Z_C, getData(e, "known"), type="ratings")


# aggregate the performance statistics
error_ICOS <- rbind(
  IBCF_N_C = calcPredictionAccuracy(p1, getData(e, "unknown")),
  IBCF_C_C = calcPredictionAccuracy(p2, getData(e, "unknown")),
  IBCF_Z_C = calcPredictionAccuracy(p3, getData(e, "unknown"))
)

kable(error_ICOS)

```

### Item-Based Collaborative Filtering: Euclidean Distance

```{r}
#train IBCF Euclidean Distance models

# non-normalized
IBCF_N_E <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = NULL, method="Euclidean"))

# centered
IBCF_C_E <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "center",method="Euclidean"))

# Z-score normalization
IBCF_Z_E <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "Z-score",method="Euclidean"))
```

```{r}
# compute predicted ratings
p1 <- predict(IBCF_N_E, getData(e, "known"), type="ratings")
p2 <- predict(IBCF_C_E, getData(e, "known"), type="ratings")
p3 <- predict(IBCF_Z_E, getData(e, "known"), type="ratings")


# aggregate the performance statistics
error_IEUC <- rbind(
  IBCF_N_E = calcPredictionAccuracy(p1, getData(e, "unknown")),
  IBCF_C_E = calcPredictionAccuracy(p2, getData(e, "unknown")),
  IBCF_Z_E = calcPredictionAccuracy(p3, getData(e, "unknown"))
)
kable(error_IEUC)
```


### Item-Based Collaborative Filtering: Pearson Correlation

```{r}
#train IBCF pearson correlation models

# non-normalized
IBCF_N_P <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = NULL, method="pearson"))

# centered
IBCF_C_P <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "center",method="pearson"))

# Z-score normalization
IBCF_Z_P <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "Z-score",method="pearson"))

```


```{r}
# compute predicted ratings
p1 <- predict(IBCF_N_P, getData(e, "known"), type="ratings")
p2 <- predict(IBCF_C_P, getData(e, "known"), type="ratings")
p3 <- predict(IBCF_Z_P, getData(e, "known"), type="ratings")

# aggregate the performance statistics
error_IPC <- rbind(
  IBCF_N_P = calcPredictionAccuracy(p1, getData(e, "unknown")),
  IBCF_C_P = calcPredictionAccuracy(p2, getData(e, "unknown")),
  IBCF_Z_P = calcPredictionAccuracy(p3, getData(e, "unknown"))
)
kable(error_IPC)

```


## SVD models

- One is Latent-Factor Collaborative Filtering Recommender
- SVD with default parameter, we call it no-para

```{r}
# Latent-Factor Collaborative Filtering Recommender SVD
latent_factor_cofi_rec_SVD <- Recommender(data=getData(e, "train"), method="SVD",parameter=list(categories=30,normalize='center',method='Pearson'))

# SVD with default parameters
no_para_svd <- Recommender(data=getData(e, "train"), method="SVD")
```


```{r}
# compute predicted ratings
p1 <- predict(latent_factor_cofi_rec_SVD, getData(e, "known"), type="ratings")
p2 <- predict(no_para_svd, getData(e, "known"), type="ratings")

# aggregate the performance statistics
error_SVD <- rbind(
  latent_factor_cofi_rec_SVD = calcPredictionAccuracy(p1, getData(e, "unknown")),
  no_para_SVD = calcPredictionAccuracy(p2, getData(e, "unknown"))
)
kable(error_SVD)

```





```{r}
# memory cleanup
rm(IBCF_N_P, IBCF_C_P, IBCF_Z_P)
```

```{r}
c_res <- data.frame(rbind(error_UCOS, error_UEUC, error_UPC, error_ICOS, error_IEUC, error_IPC, error_SVD))
c_res <- c_res[order(c_res$RMSE ),]
kable(c_res)
```


```{r}
barplot(c_res$RMSE, col = "yellow", main = "Model RMSEs", las = 2, ylab = "RMSE", horiz = FALSE, names.arg = rownames(c_res), cex.names=.8)
```





```{r}
Sys.time() - start_time
```

## Reference: 
1. Piazza:  https://piazza.com/class/kdf6l5f8bb78j?cid=868
2. Moviewlens: https://grouplens.org/datasets/movielens/
3. Dataset README: https://github.com/tonymazn/stat542/tree/main/data/README
4. Machine Learning Project – Data Science Movie Recommendation System Project in R 
   https://data-flair.training/blogs/data-science-r-movie-recommendation/
5. Movie Recommendation System: https://jeknov.shinyapps.io/movieRec/
6. Item-Based Collaborative Filtering Recommendation: 
   https://www.kaggle.com/hendraherviawan/itembased-collaborative-filter-recommendation-r
7. User-Based and Item-Based Collaborative Filtering https://rpubs.com/jt_rpubs/285729
8. Movie Recommendation System  https://jeknov.shinyapps.io/movieRec/
