---
title: "Project 4: System 2 - User-Based Collaborative Filtering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## reference: https://raw.githubusercontent.com/ChicagoBoothML/MachineLearning_Fall2015/master/Programming%20Scripts/MovieLens%20Movie%20Recommendation/R/MovieLens_MovieRecommendation.Rmd
## data: https://grouplens.org/datasets/movielens/

```{r include=FALSE}
start_time <- Sys.time()

library(recommenderlab)
library(ggplot2)
library(data.table)
library(reshape2)
```


```{r include=FALSE}
read <- function(fileName, separators) {
    data <- readLines(con <- file(fileName))
    close(con)
    records <- sapply(data, strsplit, split=separators)
    dataFrame <- data.frame(t(sapply(records,c)))
    rownames(dataFrame) <- 1: nrow(dataFrame)
    return(as.data.frame(dataFrame,stringsAsFactors = FALSE))
}


basedir ="data/train/"
movies = read(paste0(basedir,"movies.dat"), "::")
ratings = read(paste0(basedir,"ratings.dat"), "::")
users = read(paste0(basedir, "users.dat"), "::")
colnames(movies) = c('MovieID', 'title', 'genres')
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

```


```{r}
nrow(movies)
nrow(ratings)
nrow(users)

```

```{r}
#data <- parse_movielens_1m_data()
#movies <- data$movies
#users <- data$users
#ratings <- data$ratings[ , .(user_id, movie_id, rating)]
#ratings[ , `:=`(user_id = factor(user_id),
#                movie_id = factor(movie_id))]

ratings <- as(ratings, 'realRatingMatrix')
```

```{r}
train_proportion <- .5
nb_of_given_ratings_per_test_user <- 10

evaluation_scheme <- evaluationScheme(
  ratings, 
  method='split',
  train=train_proportion,
  k=1,
  given=nb_of_given_ratings_per_test_user)

evaluation_scheme
```


- Training data:
```{r}
ratings_train <- getData(evaluation_scheme, 'train')

ratings_train
```

- Test data: "known"/"given" ratings:

```{r}
ratings_test_known <- getData(evaluation_scheme, 'known')

ratings_test_known
```

- Test data: "unknown" ratings to be predicted and evaluated against:

```{r}
ratings_test_unknown <- getData(evaluation_scheme, 'unknown')

ratings_test_unknown
```

# Recommendation Models

Let's now train a number of recommendation models. The methods available in the **`recommenderlab`** package are:

```{r}
recommenderRegistry$get_entry_names()
```


The descriptions and default parameters for the methods applicable to a Real Rating Matrix are as follows:

```{r}
recommenderRegistry$get_entries(dataType='realRatingMatrix')
```



## Popularity-Based Recommender

The description and default parameters of this method in **`recommenderlab`** are as follows:

```{r}
recommenderRegistry$get_entry('POPULAR', dataType='realRatingMatrix')
```


We train a popularity-based recommender as follows:

```{r}
popular_rec <- Recommender(
  data=ratings_train,
  method='POPULAR')

popular_rec
```



## User-Based Collaborative-Filtering Recommender

User-Based Collaborative Filtering("**UBCF**") assumes that users with similar preferences will rate items similarly. Thus missing ratings for a user can be predicted by first finding a _**neighborhood**_ of similar users and then aggregate the ratings of these users to form a prediction.

The description and default parameters of this method in **`recommenderlab`** are as follows:

```{r}
recommenderRegistry$get_entry('UBCF', dataType='realRatingMatrix')
```

We train a UBCF recommender as follows:

```{r}
# User-based Collaborative Filtering Recommender
user_based_cofi_rec <- Recommender(
  data=ratings_train,
  method='UBCF',           # User-Based Collaborative Filtering
  parameter=list(
    normalize='center',    # normalizing by subtracting average rating per user;
                           # note that we don't scale by standard deviations here;
                           # we are assuming people rate on the same scale but have
                           # different biases
    method='Pearson',      # use Pearson correlation
    nn=30                  # number of Nearest Neighbors for calibration
  ))

user_based_cofi_rec
```

## Item-Based Collaborative-Filtering Recommender

Item-Based Collaborative Filtering ("**IBCF**") is a model-based approach which produces recommendations based on the relationship between items inferred from the rating matrix. The assumption behind this approach is that users will prefer items that are similar to other items they like.

The model-building step consists of calculating a similarity matrix containing all item-to-item
similarities using a given similarity measure. Popular measures are Pearson correlation and
Cosine similarity. For each item only a list of the $k$ most similar items and their similarity values are stored. The $k$ items which are most similar to item $i$ is denoted by the set $S(i)$ which can be seen as the neighborhood of size $k$ of the item. Retaining only $k$ similarities per item improves the space and time complexity significantly but potentially sacrifices some recommendation quality.

The description and default parameters of this method in **`recommenderlab`** are as follows:

```{r}
recommenderRegistry$get_entry('IBCF', dataType='realRatingMatrix')
```

We train a IBCF recommender as follows:

```{r}
# Item-based Collaborative Filtering Recommender
item_based_cofi_rec <- Recommender(
  data=ratings_train,
  method='IBCF',           # Item-Based Collaborative Filtering
  parameter=list(
    normalize='center',    # normalizing by subtracting average rating per user;
                           # note that we don't scale by standard deviations here;
                           # we are assuming people rate on the same scale but have
                           # different biases
    method='Pearson',      # use Pearson correlation
    k=100                  # number of Nearest Neighbors for calibration
  ))

item_based_cofi_rec
```











```{r}
# Evaluation:
#evaluation_scheme <- evaluationScheme(ratingmat, 
#                                      method="cross-validation", 
#                                      k=5, given=3, 
#                                      goodRating=5) #k=5 meaning a 5-fold cross validation. given=3 meaning a Given-3 protocol
#evaluation_results <- evaluate(evaluation_scheme, 
#                               method="UBCF", 
#                               n=c(1,3,5,10,15,20))
#eval_results <- getConfusionMatrix(evaluation_results)[[1]]


```

## Latent-Factor Collaborative-Filtering Recommender

_**Note**: as of the date of writing, the **`recommenderlab`** package's default implementation of the Latent-Factor Collaborative Filtering / Singular Value Decomposition (SVD) method has a bug and produces dodgy performances. The code in this section is commented out for now._

This approache uses Singular-Value Decomposition (SVD) to factor the Rating Matrix into a product of user-feature and item-feature matrices.

The description and default parameters of this method in **`recommenderlab`** are as follows:

```{r}
recommenderRegistry$get_entry('SVD', dataType='realRatingMatrix')
```

We train a Latent-Factor CF recommender as follows:

```{r}
# Latent-Factor Collaborative Filtering Recommender
# with matrix factorization by Singular-Value Decomposition (SVD)
 latent_factor_cofi_rec <- Recommender(
   data=ratings_train,
   method='SVD',            # Item-Based Collaborative Filtering
   parameter=list(
     categories=30,         # number of latent factors
     normalize='center',    # normalizing by subtracting average rating per user;
                           # note that we don't scale by standard deviations here;
                           # we are assuming people rate on the same scale but have
                           # different biases
     method='Pearson'       # use Pearson correlation
   ))

 latent_factor_cofi_rec
```

# Model Evaluation

Now, we make predictions on the Test set and and evaluate these recommenders' OOS performances:

```{r}
popular_rec_pred <- predict(
  popular_rec,
  ratings_test_known,
  type='ratings')

popular_rec_pred_acc <- calcPredictionAccuracy(
  popular_rec_pred,
  ratings_test_unknown)

popular_rec_pred_acc
```

```{r}
#user_based_cofi_rec_pred <- predict(
#  user_based_cofi_rec,
#  ratings_test_known,
#  type='ratings')

#user_based_cofi_rec_pred_acc <- calcPredictionAccuracy(
#  user_based_cofi_rec_pred,
#  ratings_test_unknown)

#user_based_cofi_rec_pred_acc
```

```{r}
item_based_cofi_rec_pred <- predict(
  item_based_cofi_rec,
  ratings_test_known,
  type='ratings')

item_based_cofi_rec_pred_acc <- calcPredictionAccuracy(
  item_based_cofi_rec_pred,
  ratings_test_unknown)

item_based_cofi_rec_pred_acc
```



```{r}
latent_factor_cofi_rec_pred <- predict(
  latent_factor_cofi_rec,
  ratings_test_known,
  type='ratings')

latent_factor_cofi_rec_pred_acc <- calcPredictionAccuracy(
  latent_factor_cofi_rec_pred,
  ratings_test_unknown)

latent_factor_cofi_rec_pred_acc
```






```{r}
Sys.time() - start_time
```

