---
title: "Project 4: Movie Recommendation Report by ZM11"
author: Zhouning Ma (ZM11)
date: '09/07/2020'
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan

fontfamily: mathpazo
fontfamilyoptions: sc, osf
fontsize: 11pt
linkcolor: blue
---

***



```{r include=FALSE}
start_time <- Sys.time()


library(recommenderlab)
library(ggplot2)
library(data.table)
library(reshape2)
library(knitr)

set.seed(4486)
```


### Project Overview:
In this project, we develop content based recommendation system(System I) and collaborative recommendation system(System II).

* **System I:** recommendation based on genres. We know user's favorite genre, we provide two recommendation schemes: 
   + __Method I:__  highly-rate movies - select the 4-point or more review
   + __Method II:__ most trendy movies - the newest movie(here is year 2000)

* **System II:** collaborative recommendation system. we provided three collaborative recommendation algorithms: 
   + __User-based (UBCF):__ we assume that similar users will have similar taste. UBCF uses the logic and recommends items by finding similar users to the user;
   + __Item-based (IBCF):__ The similarities between different items in the dataset are calculated by using one of a number of similarity measures, and then these similarity values are used to predict;
   + __Singular value decomposition (SVD):__ It uses a matrix structure where each row represents a user, and each column represents an item. The elements of this matrix are the ratings that are given to items by users.


```{r include=FALSE}
read <- function(fileName, separators) {
    data <- readLines(con <- file(fileName))
    close(con)
    records <- sapply(data, strsplit, split=separators)
    dataFrame <- data.frame(t(sapply(records,c)))
    rownames(dataFrame) <- 1: nrow(dataFrame)
    return(as.data.frame(dataFrame,stringsAsFactors = FALSE))
}


basedir ="data/test/"
movies = read(paste0(basedir,"movies.dat"), "::")
ratings = read(paste0(basedir,"ratings.dat"), "::")
users = read(paste0(basedir, "users.dat"), "::")
colnames(movies) = c('MovieID', 'title', 'genres')
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

```


### Dataset: 
The dataset contains about 1 million ratings and the details as following:

* Number of movies: $`r nrow(movies)`$
* Number of ratings: $`r nrow(ratings)`$
* Number of users: $`r nrow(users)`$

The dataset README(3) shows:
Genres are pipe-separated and are selected from the following genres:

* Action
* Adventure
* Animation
* Children's
* Comedy
* Crime
* Documentary
* Drama
* Fantasy
* Film-Noir
- Horror
- Musical
- Mystery
- Romance
- Sci-Fi
- Thriller
- War
- Western
	




## Create realRatingMatrix

```{r}
ratings <- as(ratings, 'realRatingMatrix')
```


## Training and Testing Data
We split the data into 80% training and 20% testing. 

```{r}
train_proportion <- 0.9
to_keep <- 15  ## given 15 items
threshold <- 0 ## ratings above 0 as the cutoff point

# split the data into the training and the test set:
e <- evaluationScheme(ratings, method="split", train=train_proportion, k=10, given=to_keep, goodRating=threshold)
#e <- evaluationScheme(ratings, method="split", train=train_proportion, given=to_keep, goodRating=threshold)
e

```



```{r}
# Latent-Factor Collaborative Filtering Recommender
# with matrix factorization by Singular-Value Decomposition (SVD)
latent_factor_cofi_rec_SVD <- Recommender(data=getData(e, "train"), method="SVD",parameter=list(categories=30,normalize='center',method='Pearson'))
no_para_svd <- Recommender(data=getData(e, "train"), method="SVD")
```



```{r}
# compute predicted ratings
p1 <- predict(latent_factor_cofi_rec_SVD, getData(e, "known"), type="ratings")
p2 <- predict(no_para_svd, getData(e, "known"), type="ratings")

# aggregate the performance statistics
error_SVD <- rbind(
  latent_factor_cofi_rec_SVD = calcPredictionAccuracy(p1, getData(e, "unknown")),
  SVD_C_P = calcPredictionAccuracy(p2, getData(e, "unknown"))
)
kable(error_SVD)

```






```{r}
Sys.time() - start_time
```

## Reference: 
1. Piazza:  https://piazza.com/class/kdf6l5f8bb78j?cid=868
2. Moviewlens: https://grouplens.org/datasets/movielens/
3. Dataset README: https://github.com/tonymazn/stat542/tree/main/data/README
4. Machine Learning Project – Data Science Movie Recommendation System Project in R https://data-flair.training/blogs/data-science-r-movie-recommendation/
5. Movie Recommendation System: https://jeknov.shinyapps.io/movieRec/
6. Item-Based Collaborative Filtering Recommendation: https://www.kaggle.com/hendraherviawan/itembased-collaborative-filter-recommendation-r
7. User-Based and Item-Based Collaborative Filtering https://rpubs.com/jt_rpubs/285729
8. Movie Recommendation System  https://jeknov.shinyapps.io/movieRec/
