---
title: "Project 4: Movie Recommendation Report"
author: Zhouning Ma (ZM11)
date: '09/07/2020'
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan

fontfamily: mathpazo
fontfamilyoptions: sc, osf
fontsize: 11pt
linkcolor: blue
---

***



```{r include=FALSE}
start_time <- Sys.time()

library(recommenderlab)
library(ggplot2)
library(data.table)
library(reshape2)
library(knitr)
```


### Project Overview
In this project, we develop content based recommendation system(System I) and collaborative recommendation system(System II) (1).

* **System I:** recommendation based on genres. We know user's favorite genre, we provide two recommendation schemes: 
   + __Method I:__  highly-rate movies - select the 4-point or more review
   + __Method II:__ most trendy movies - the newest movie(here is year 2000)

* **System II:** collaborative recommendation system. we provided three collaborative recommendation algorithms: 
   + __User-based (UBCF):__ we assume that similar users will have similar taste. UBCF uses the logic and recommends items by finding similar users to the user;
   + __Item-based (IBCF):__ The similarities between different items in the dataset are calculated by using one of a number of similarity measures, and then these similarity values are used to predict;
   + __Singular value decomposition (SVD):__ It uses a matrix structure where each row represents a user, and each column represents an item. The elements of this matrix are the ratings that are given to items by users.


```{r include=FALSE}
read <- function(fileName, separators) {
    data <- readLines(con <- file(fileName))
    close(con)
    records <- sapply(data, strsplit, split=separators)
    dataFrame <- data.frame(t(sapply(records,c)))
    rownames(dataFrame) <- 1: nrow(dataFrame)
    return(as.data.frame(dataFrame,stringsAsFactors = FALSE))
}


basedir ="data/test/"
movies = read(paste0(basedir,"movies.dat"), "::")
ratings = read(paste0(basedir,"ratings.dat"), "::")
users = read(paste0(basedir, "users.dat"), "::")
colnames(movies) = c('MovieID', 'title', 'genres')
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

moviesList <- read(paste0(basedir,"aggr.dat"), "::")
movies_clean <- read(paste0(basedir,"movies_clean.dat"), "::")

colnames(moviesList) = c( 'MovieID', 'AveRating', 'title', 'genres')
colnames(movies_clean) = c("MovieID", "title", "year", 'genres')
genre_list <- c("Action", "Adventure", "Animation", "Children", 
                "Comedy", "Crime","Documentary", "Drama", "Fantasy",
                "Film-Noir", "Horror", "Musical", "Mystery","Romance",
                "Sci-Fi", "Thriller", "War", "Western")
```


### Dataset
The dataset(2) contains about 1 million ratings and the details as following:

* Number of movies: $`r nrow(movies)`$
* Number of ratings: $`r nrow(ratings)`$
* Number of users: $`r nrow(users)`$

The dataset README(3) shows:
Genres are pipe-separated and are selected from the following genres:

* Action
* Adventure
* Animation
* Children's
* Comedy
* Crime
* Documentary
* Drama
* Fantasy
* Film-Noir
- Horror
- Musical
- Mystery
- Romance
- Sci-Fi
- Thriller
- War
- Western
	



## System I

system I is content based recommendation system, let's say we know the user's favorite genres, we will recommend movies based on genres. 

### Method I: highly-rate movies

  We calculate the average point of each movie and randomly pick N movies which average point is great than 3-point(4-point or 5-point) in the genres. 


```{r}
  # Method 1:
  numberofmovierecommend = 5

  input_genre1 = genre_list[1]
  input_genre2 = genre_list[2]
  input_genre3 = genre_list[3]

  systemresult = subset(moviesList,AveRating>=4 & (grepl(input_genre1, genres, fixed = TRUE) | grepl(input_genre2, genres, fixed = TRUE) | grepl(input_genre3, genres, fixed = TRUE)) )
  if (nrow(systemresult) < numberofmovierecommend){
     systemresult = subset(moviesList, grepl(input_genre1, genres, fixed = TRUE) | grepl(input_genre2, genres, fixed = TRUE) | grepl(input_genre3, genres, fixed = TRUE))
  }

  systemresult = systemresult[sample(nrow(systemresult), ifelse(nrow(systemresult)>=numberofmovierecommend,numberofmovierecommend,nrow(systemresult))),]
  systemresult


```
  
  
  
### Method II:most trendy movies

  The newest movies are year 2000 in the dataset, we will randomly pick N movies from year 2000 in the genres, Unfortunately sometimes there are not enough movies
  in year 2000 for each genres, for example, it is empty in genres "Film-Noir" of year 2000, we have to try the movies of year 1999.
  

```{r}
  # Method 2:
  numberofmovierecommend = 5
  trendyYear = 2000

  input_genre1 = genre_list[1]
  input_genre2 = genre_list[2]
  input_genre3 = genre_list[3]


  systemresult = subset(movies_clean, year >= trendyYear & (grepl(input_genre1, genres, fixed = TRUE) | grepl(input_genre2, genres, fixed = TRUE) | grepl(input_genre3, genres, fixed = TRUE)) )
  if (nrow(systemresult) < numberofmovierecommend){
      systemresult = subset(movies_clean, year >= trendyYear - 1 & (grepl(input_genre1, genres, fixed = TRUE) | grepl(input_genre2, genres, fixed = TRUE) | grepl(input_genre3, genres, fixed = TRUE)) )
  }
  systemresult = systemresult[sample(nrow(systemresult), ifelse(nrow(systemresult)>=numberofmovierecommend,numberofmovierecommend,nrow(systemresult))),]
  systemresult
```



## Evaluation result
It looks like the SVDs' performance(prediction accuracy) are better than User-based and Item-based Collaborative Filtering. 


## Save pre-trained models
It takes time to train a model with large data, to save time for the App, we will store the model object to  .rds format, when the system starts, 
it will load the .rds to the model object. it will be very fast. We save all the models we tested, it is around 20 models.

## Movie Recommendation App

  * it is a shiny app with System I and System II algorithms
  * System I
    + the app needs to take the input from users on their favorite genres, and display the result
    + system will provide two methods for System I (you are able to select any of them by the setting)
  
 * System II 
   + the app need the users to rate many movies as possible;
   + system will discover movies for the users by User-based or Item-based Collaborative Filtering models or SVD models

 * Setting
   + select the algorithm for System I
   + select the algorithm for System II


## Deploy the App
it is a huge challenge to deploy to Shiny server, it doesn't support the dataset like us(1M rows), but we found a way finally to make it successfully, 
we are able to running the App on the full dataset, but it is not for production, it is ok for the proof of concept. We found it is stable when we work
RStudio on windows environment, why don't we use this one to demonstrate? Once we are able to change the web server port to 80, then we are able to host.
We run the following code on RStudio console:

**runApp('app', host = "0.0.0.0", port = 80)**


anther thing we need to do is the server will shutdown in 1 hour, we need to ping(wget can do this job) it every 10 minutes.



## Conclusion

Multiple models were compared on the dataset, the SVD model is the top one performance. System II Collaborative Filtering is an improvement over the 
System I (content based recommendation system). They can take many classes of input and provide similarity ranking based algorithms to provide the user
with accurate results. These recommendation systems have evolved over time and have incorporated many advanced techniques to provide users with the content they want.


> Conten-Base (9)

* Pros: 
  + do not use user data, start recommending on day 1
  + can recommend new and unpopular items
  + can recommend to users with unique taste
  + easier to interpret
      
* Cons
  + cannot recommend outside the user's profile
  + recommend substitute not compliment
  + finding appropriate features is diffcult
      
> Collaborative Filtering (CF)

* Pros:
  + no need to define features
  + can recommend outside the user's profile
  + recommend substitute not compliment

* Cons: 
  + need enough user data to start, cannot operate on day 1
  + cannot recommnend new, unrated items
  + tend to recommend popular items, against the purpose of the personalized RS
  + cold start problem for new users/items
      
The item-base CF performs better in pratice: easier to find similar items, but difficult to find similar people.










