---
title: "Project 4: System 2 - User-Based Collaborative Filtering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## reference: https://raw.githubusercontent.com/ChicagoBoothML/MachineLearning_Fall2015/master/Programming%20Scripts/MovieLens%20Movie%20Recommendation/R/MovieLens_MovieRecommendation.Rmd
## data: https://grouplens.org/datasets/movielens/

```{r include=FALSE}
start_time <- Sys.time()

library(recommenderlab)
library(ggplot2)
library(data.table)
library(reshape2)
library(knitr)

```


```{r include=FALSE}
read <- function(fileName, separators) {
    data <- readLines(con <- file(fileName))
    close(con)
    records <- sapply(data, strsplit, split=separators)
    dataFrame <- data.frame(t(sapply(records,c)))
    rownames(dataFrame) <- 1: nrow(dataFrame)
    return(as.data.frame(dataFrame,stringsAsFactors = FALSE))
}


basedir ="data/"
movies = read(paste0(basedir,"movies.dat"), "::")
ratings = read(paste0(basedir,"ratings.dat"), "::")
users = read(paste0(basedir, "users.dat"), "::")
colnames(movies) = c('MovieID', 'title', 'genres')
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

```


```{r}
nrow(movies)
nrow(ratings)
nrow(users)

```

```{r}
#data <- parse_movielens_1m_data()
#movies <- data$movies
#users <- data$users
#ratings <- data$ratings[ , .(user_id, movie_id, rating)]
#ratings[ , `:=`(user_id = factor(user_id),
#                movie_id = factor(movie_id))]

ratings <- as(ratings, 'realRatingMatrix')
```

```{r}
train_proportion <- 1
nb_of_given_ratings_per_test_user <- 10

#evaluation_scheme <- evaluationScheme(ratings, method='split', train=train_proportion, k=1, given=nb_of_given_ratings_per_test_user)
# split the data into the training and the test set:
e <- evaluationScheme(ratings, method="split", train=train_proportion, given=15, goodRating=0)
e
```


## User-Based Collaborative Filtering: Cosine Similarity
- UBCF_N_C : The raw data is used with no normalization applied;

- UBCF_C_C: Data are normalized using centering.

- UBCF_Z_C: Z-score normalization is applied to the data;

```{r}
#train UBCF cosine similarity models

# non-normalized
UBCF_N_C <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = NULL, method="Cosine"))
saveRDS(UBCF_N_C, file = "model/UBCF_N_C_model.rds")
reloadModel = readRDS("model/UBCF_N_C_model.rds")



# centered
UBCF_C_C <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "center",method="Cosine"))
saveRDS(UBCF_C_C, file = "model/UBCF_C_C_model.rds")
reloadModel = readRDS("model/UBCF_C_C_model.rds")


# Z-score normalization
UBCF_Z_C <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "Z-score",method="Cosine"))
saveRDS(UBCF_Z_C, file = "model/UBCF_Z_C_model.rds")
reloadModel = readRDS("model/UBCF_Z_C_model.rds")


```


## User-Based Collaborative Filtering: Euclidean Distance
```{r}
#train UBCF Euclidean Distance models

# non-normalized
UBCF_N_E <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = NULL, method="Euclidean"))
saveRDS(UBCF_N_E, file = "model/UBCF_N_E_model.rds")
reloadModel = readRDS("model/UBCF_N_E_model.rds")


# centered
UBCF_C_E <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "center",method="Euclidean"))
saveRDS(UBCF_C_E, file = "model/UBCF_C_E_model.rds")
reloadModel = readRDS("model/UBCF_C_E_model.rds")


# Z-score normalization
UBCF_Z_E <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "Z-score",method="Euclidean"))
saveRDS(UBCF_Z_E, file = "model/UBCF_Z_E_model.rds")
reloadModel = readRDS("model/UBCF_Z_E_model.rds")

```


# User-Based Collaborative Filtering: Pearson Correlation
```{r}
#train UBCF pearson correlation models

# non-normalized
UBCF_N_P <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = NULL, method="pearson"))
saveRDS(UBCF_N_P, file = "model/UBCF_N_P_model.rds")
reloadModel = readRDS("model/UBCF_N_P_model.rds")


# centered
UBCF_C_P <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "center",method="pearson"))
saveRDS(UBCF_C_P, file = "model/UBCF_C_P_model.rds")
reloadModel = readRDS("model/UBCF_C_P_model.rds")

# Z-score normalization
UBCF_Z_P <- Recommender(getData(e, "train"), "UBCF", param=list(normalize = "Z-score",method="pearson"))
saveRDS(UBCF_Z_P, file = "model/UBCF_Z_P_model.rds")
reloadModel = readRDS("model/UBCF_Z_P_model.rds")

```



# Item-Based Collaborative Filtering: Cosine Similarity
```{r}
#train IBCF cosine similarity models

# non-normalized
IBCF_N_C <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = NULL, method="Cosine"))
saveRDS(IBCF_N_C, file = "model/IBCF_N_C_model.rds")
reloadModel = readRDS("model/IBCF_N_C_model.rds")

# centered
IBCF_C_C <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "center",method="Cosine"))
saveRDS(IBCF_C_C, file = "model/IBCF_C_C_model.rds")
reloadModel = readRDS("model/IBCF_C_C_model.rds")

# Z-score normalization
IBCF_Z_C <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "Z-score",method="Cosine"))
saveRDS(IBCF_Z_C, file = "model/IBCF_Z_C_model.rds")
reloadModel = readRDS("model/IBCF_Z_C_model.rds")

```



# Item-Based Collaborative Filtering: Euclidean Distance
```{r}
#train IBCF Euclidean Distance models

# non-normalized
IBCF_N_E <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = NULL, method="Euclidean"))
saveRDS(IBCF_N_E, file = "model/IBCF_N_E_model.rds")
reloadModel = readRDS("model/IBCF_N_E_model.rds")

# centered
IBCF_C_E <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "center",method="Euclidean"))
saveRDS(IBCF_C_E, file = "model/IBCF_C_E_model.rds")
reloadModel = readRDS("model/IBCF_C_E_model.rds")

# Z-score normalization
IBCF_Z_E <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "Z-score",method="Euclidean"))
saveRDS(IBCF_Z_E, file = "model/IBCF_Z_E_model.rds")
reloadModel = readRDS("model/IBCF_Z_E_model.rds")

```


# Item-Based Collaborative Filtering: Pearson Correlation
```{r}
#train IBCF pearson correlation models

# non-normalized
IBCF_N_P <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = NULL, method="pearson"))
saveRDS(IBCF_N_P, file = "model/IBCF_N_P_model.rds")
reloadModel = readRDS("model/IBCF_N_P_model.rds")

# centered
IBCF_C_P <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "center",method="pearson"))
saveRDS(IBCF_C_P, file = "model/IBCF_C_P_model.rds")
reloadModel = readRDS("model/IBCF_C_P_model.rds")

# Z-score normalization
IBCF_Z_P <- Recommender(getData(e, "train"), "IBCF", param=list(normalize = "Z-score",method="pearson"))
saveRDS(IBCF_Z_P, file = "model/IBCF_Z_P_model.rds")
reloadModel = readRDS("model/IBCF_Z_P_model.rds")
```



# SVD
```{r}
recommenderRegistry$get_entry('SVD', dataType='realRatingMatrix')
```

```{r}
# Latent-Factor Collaborative Filtering Recommender
# with matrix factorization by Singular-Value Decomposition (SVD)
 latent_factor_cofi_rec_SVD <- Recommender(
   data=getData(e, "train"),
   method='SVD',            # Item-Based Collaborative Filtering
   parameter=list(
     categories=30,         # number of latent factors
     normalize='center',    # normalizing by subtracting average rating per user;
                            # note that we don't scale by standard deviations here;
                            # we are assuming people rate on the same scale but have
                            # different biases
     method='Pearson'       # use Pearson correlation
   ))

 saveRDS(latent_factor_cofi_rec_SVD, file = "model/latent_factor_cofi_rec_SVD_model.rds")
 reloadModel = readRDS("model/latent_factor_cofi_rec_SVD_model.rds")

 svd <- Recommender(data = getData(e, "train"), method = "SVD")
 saveRDS(svd, file = "model/No_Para_SVD_model.rds")
 reloadModel = readRDS("model/No_Para_SVD_model.rds")
 
```


```{r}
Sys.time() - start_time
```

